---
title: "SkeletalVis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SkeletalVis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Getting started

### Installation



### Loading the package
Once skeletavis is installed the package needs to be loaded into a user’s R session.
```{r setup}
library(SkeletalVis)
library(feather)
```

### Accessing the skeletalvis dataset
First we need to download the gene expression data and metadata. The data is stored as a set of feather files, which allow on disk storage of large datsets containing the foldchanges, pvalues and metadata files describing each dataset. Advanced users can simplify explore these files using R tools but for conveience we provide functions that allow common  tasks to be performed with minimal knowledge of R.
```{r}
skeletalvis_folder <- download_skeletalvis()
```

## Exploring skeletalvis
We can explore SkeletalVis in multiple ways, finding genes of interest, finding an experimental of interest or finding experiments that have gene expression profile similar to a query.

We can view the differential expression of a set of genes of interest with the `get_gene_data` function.
```{r}
gene_results <- get_gene_fold_changes(skeletalvis, c("SOX9","COL2A1"), pvalues=TRUE)
```
### Find a dataset
We can search for an invidiual dataset of interest by querying the metadata table. For example we might be interested in a particular species and disease.
```{r}
experiments <- search_skeletalvis(skeletalvis, c("Mouse","Osteoarthritis"))
```

Alterively we can intereactive search the table of metadata and to find experiments of interest
```{r}
browse_skeletalvis(skeletalvis)
```

We can get the fold changes and pValues (if applicable) for an individual dataset by search by skeletalvis ID. This ID is the GEO accession number and the comparison number (if multiple comparisons are made within the same dataset e.g mut1 vs control, mut2 vs control).

```{r}
experiment_results <- get_experiment(skeletalvis, "GSE107363_1")
```

#### Volcano plots
We can plot the differential expression of an individual volcano plot with either plotly or ggplot2 for interactive and static plots respectively.

```{r}
volcano_plot(experiment_results, numberGenes=10, interactive = FALSE)
```

```{r}
volcano_plot(experiment_results, numberGenes=10, interactive = TRUE)
```
We can also plot genes of interest on the volcano plot
```{r}
volcano_plot(experiment_results, numberGenes=10, selectedGenes=c("SOX9","COL2A1","ACAN"))
```

## Finding similar datasets
We can find similar datasets using either cosine similarity (measure the correlation of fold changes across all the SkeletalVis data). As input we need a dataframe of human gene symbols and fold changes.
```{r}
# Get a recent dataset to compare
query <- data.frame(ID=c("COL2A1","SOX9","COL1A1","ACAN","MMP3","MMP13"),
                      foldChange=c(1.3,1.2,0.5,-3,3,0.85)
                      )

# Get the similarity of this experiment against the skeletalvis databases
similarityTable <- experiment_similarity(query,skeletalvis)
  
```
We can view the results of the search, the table is sorted by zscore so the most similar datasets are at the top.
```{r}
# Look at the top few rows
head(similarityTable)
```
We can view the most dissimilar datasets by looking at the bottom of the table.
```{r}
# Look at the bottom few rows
tail(similarityTable)
```

### Plotting the similarities
We can plot the most similar datasets by plotting the similarity zscore against the rank of the dataset and label the top few datasets.
```{r}
# Plot the similarity results and label the top 5 experiments
plot_similarity(similarityTable, top_n=5)
```



